Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-05-02 doc/manuscript_hkk_drought_sensitivity.Rmd
results and part of discussion from sean

In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including in precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum temperature (max anomaly +1.5 SD), and VPD (max anomaly +2 SD).
1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought.

@@

Individual tree responses to drought were not correlated with each other (Fig S6).

TWI had no significant on sensitivity in 2010 but a positive effect in 2015. 
Specifically, across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

@@

<!---CONCLUSIONS---->
We show that, at least in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found lower-than-average growth in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`). 
Comparing across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
While trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
In other words, water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees (with likely deeper roots) fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`). 
Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce enormous heterogeneity in individual and species drought responses of tropical trees.
This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diveristy and ecosystem stability.

<!----Limitations in filling gap/ comparing only two droughts---->
While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size. 
<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments. 
Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.  -->
<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig. `r fig_droughts`).
Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
While our finding that a wet-season drought (2015) had greater impact on woody growth than a dry-season drought (2010) calls into question the generality of previous findings based on tree-ring chronologies from HKK [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation, care must be taken in interpretation given that the 2015 drought was more anomalous on multi-month time scales (Fig. `r fig_droughts`).
Rigorous generalization on tropical tree growth responses to any specific drought characteristic would require annual growth records from numerous species over numerous droughts, but no such data set currently exists.
Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
As tropical dendrometer band records continues to grow, so will the potential to better understand how specific drought characteristics interact with species traits and microenironment to shape growth respones.

>>>>

In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
<!-- make sure my adjustments above in details are correct ... I think we should present a precise description  -->
1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought.

@@

The test of whether individual trees responded similarly to the two drought events showed no correlation (Fig S6).

TWI showed no significant effect on sensitivity in 2010, but did show a positive effect in 2015. 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

@@

<!---CONCLUSIONS---->
We show that, in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`). 
<!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diveristy and ecosystem stability.

<!----Limitations in filling gap/ comparing only two droughts---->
While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size. 
<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments. 
Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.  -->
<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig. `r fig_droughts`). This is a central finding of our study. 
<!-- Quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.  I don't understand this, as we have done this, no? -->
However, our finding that a wet-season drought in 2015 had greater impact on woody growth than a dry-season drought in 2010 questions the generality of previous findings. Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation. But  given that the 2015 drought was more anomalous on multi-month time scales (Fig. `r fig_droughts`) does not align with this assumption.
Generalization of tropical tree growth responses to any specific drought seems to require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) from species that might span importance and drought strategies over different timings and intensities of droughts. Advancing efforts to collect this level of data is critical to our assessment and prediction of drought responses of tropical forets to changing precipitation patterns.
<!-- [is this necessary?]  Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
As tropical dendrometer band records continues to grow, so will the potential to better understand how specific drought characteristics interact with species traits and microenironment to shape growth respones.  -->
